Data Analysis of Mental Health Trends from 2014 to 2023 in Toronto

### 2024-01-25 outputs/paper/paper.qmd
fixed some typos and added one reference

---
title: "Data Analysis of Mental Health Trends from 2014 to 2023 in Toronto"
author: "Mary Cheng"
thanks: "Code and data are avaliable at:https://github.com/marycx/Analysis-of-Mental-Health-Situation-from-2014-to-2023-in-Toronto.git"
date: today
date-format: long
abstract: "In today's society, acknowledging the importance of mental health and the growing awareness surrounding it has increased more and more. This paper explores the 'Mental Health Act (MHA) Apprehension' data from opendatatoronto. The study analyzes trends in mental health situations and their correlation with those factors. Findings illustrate that number of MHA apprehensions is at its highest in 2021. It also indicates there are higher MHA apprehensions among males. Also, the age group '25 to 34' shows the highest occurrences of MHA apprehensions. Additionally, individuals living in apartments are more likely to face MHA apprehension. The paper provides insights into the relationships among mental health conditions, age, gender, and living arrangements, offering a concise overview of the situation."
format: pdf
toc: true
bibliography: references.bib
---

## Introduction

@@

In this paper, we use R[@citeR] to analyze a dataset from opendatatoronto[@citeODT] on the occurrences of MHA apprehension in Toronto from 2014 to 2023[@citeMentalData] in order to get a better idea on the development of mental health situation and what factors are influencing individuals' mental health. This report first looks at the overall trend of mental health situation across 2014 to 2023 after conducting data cleaning and pre-processing. It then attempts to analyze the intricate relationships between mental health conditions and demographic factors such as age, sex, and premise type, using graphs and tables. It is shown that the occurrences of male MHA apprehension are higher than female MHA apprehension in each year between 2014 to 2023. Also, from various age groups, age group "25 to 34" has the highest occurrences of MHA apprehensions compared to other age groups. Moreover, it is found that the number of individuals living in "Apartment" has a higher percentage of getting MHA apprehension compared to people living in other places. However, this may be slightly biased since majority of people in Toronto lives in apartments due to the packed city.

@@

## Data

@@

### Data Pre-processing

@@

### Data Overview

@@

@tbl-sampledata provides some sample data entries in the refined dataset. It can be observed that the post-processing data comprises four columns: "Occurred Year," "Sex," "Age Group," and "Premises Type."

@tbl-sum_occurences_new shows the total number of MHA apprehensions across 2014 to 2023. A bar plot is created to better visualize the general trend of the dataset. It can be seen in @fig-total which illustrates a gradual rise in the occurrences of MHA apprehensions from 2014 to 2021, peaking in 2021 (13219 occurrences). Moreover, a slight decrease is observed from 2021 to 2023. Some implications can be inferred from this trend. The reason why 2021 has the most cases of MHA apprehensions might be due to COVID-19 pandemic which was prevailing at that time. COVID-19 brought multifaceted challenges to people, for example, the economic uncertainty, lose of loved ones, prolonged social isolation, and so on. To some extent, these challenges all contributed to the increasing number of mental health problems.

@@

### Relationship between Sex and Mental Health

@@

### Relationship between Age and Mental Health

@@

@fig-age_year_each illustrates the relationship between the age group and the number of MHA apprehension in each year more clearly. It can be observed that the number of MHA apprehensions in age group "25 to 34" is always at the highest compared to in other age groups. Also, age group "65+" always has the lowest number of MHA apprehensions. For age group "18 to 24", "25 to 34", "35 to 44" and "45 to 54", the number of MHA apprehensions increases from 2014 to 2021, reaching its peak at 2021, and then gradually decreases from 2021 to 2023. For age group "65+", the number of MHA apprehensions has a consistent increasing trend from 2014 to 2023.

@@

### Relationship between Premise Types and Mental Health

@@

As illustrated in @fig-premise_type, the distribution of the number of MHA apprehensions across various premise types is presented as percentages. It can be seen that "Apartment" emerges as the predominant category in premise types. In total, 40.9% of MHA apprehension of individuals live in "Apartment". MHA apprehensions of individuals living in "House" has the second largest percentage (17.8%) compared to others. The number of MHA apprehensions of individuals living in "Outside" and "Other" are approximately the same, which are 16.4% and 16.7% respectively. The remaining 8.2% includes individuals living in "Commercial", "Transit", "Educational" buildings, which do not have significant effects in the context of MHA apprehensions.

@@

## Reference

>>>>

---
title: "Data Analysis of Mental Health Trends from 2014 to 2023 in Toronto"
author: "Mary Cheng"
thanks: "Code and data are avaliable at:https://github.com/marycx/Analysis-of-Mental-Health-Situation-from-2014-to-2023-in-Toronto.git"
date: today
date-format: long
abstract: "In today's society, acknowledging the importance of mental health and the growing awareness surrounding it has increased more and more. This paper explores the 'Mental Health Act (MHA) Apprehension' data from opendatatoronto. The study analyzes trends in mental health situations and their correlation with those factors. Findings illustrate that number of MHA apprehensions is at its highest in 2021. It also indicates there are higher MHA apprehensions among males. The age group '25 to 34' shows the highest occurrences of MHA apprehensions. Additionally, individuals living in apartments are more likely to face MHA apprehension. The paper provides insights into the relationships among mental health conditions, age, gender, and living arrangements, offering a concise overview of the situation."
format: pdf
toc: true
bibliography: references.bib
---

## 1 Introduction

@@

In this paper, we use R[@citeR] to analyze a dataset from opendatatoronto[@citeODT] on the occurrences of MHA apprehension in Toronto from 2014 to 2023[@citeMentalData] in order to get a better idea on the development of mental health situation and what factors are influencing individuals' mental health. This report first looks at the overall trend of mental health situation across 2014 to 2023 after conducting data cleaning and pre-processing. It then attempts to analyze the intricate relationships between mental health conditions and demographic factors such as age, sex, and premise type, using graphs and tables. It is shown that the occurrences of male MHA apprehension are higher than female MHA apprehension in each year between 2014 to 2023. Also, age group "25 to 34" has the highest occurrences of MHA apprehensions compared to other age groups. Moreover, it is found that the number of individuals living in "Apartment" has a higher percentage of getting MHA apprehension compared to people living in other places. However, this may be slightly biased since majority of people in Toronto lives in apartments due to the packed city.

@@

## 2 Data

@@

### 2.1 Data Pre-processing

@@

### 2.2 Data Overview

@@

@tbl-sampledata provides some sample data entries in the refined dataset. It can be observed that the post-processing data comprises four columns: "Occurred Year," "Sex," "Age Group," and "Premises Type".

@tbl-sum_occurences_new shows the total number of MHA apprehensions across 2014 to 2023. A bar plot is created to better visualize the general trend of the dataset. It can be seen in @fig-total which illustrates a gradual rise in the occurrences of MHA apprehensions from 2014 to 2021, peaking in 2021 (13219 occurrences). Moreover, a slight decrease is observed from 2021 to 2023. Some implications can be inferred from this trend. The reason why 2021 has the most cases of MHA apprehensions might be due to COVID-19 pandemic which was prevailing at that time. COVID-19 brought multifaceted challenges to people, for example, the economic uncertainty, lose of loved ones, prolonged social isolation, and so on. To some extent, these challenges all contributed to the increasing number of mental health problems [@citecovid].

@@

### 2.3 Relationship between Sex and Mental Health

@@

### 2.4 Relationship between Age and Mental Health

@@

@fig-age_year_each illustrates the relationship between the age group and the number of MHA apprehension in each year more clearly. It can be observed that the number of MHA apprehensions in age group "25 to 34" is always at the highest compared to in other age groups. This is possibly due to various societal and economic pressure for people in this age group. Also, age group "65+" always has the lowest number of MHA apprehensions. For age group "18 to 24", "25 to 34", "35 to 44" and "45 to 54", the number of MHA apprehensions increases from 2014 to 2021, reaching its peak at 2021, and then gradually decreases from 2021 to 2023. For age group "65+", the number of MHA apprehensions has a consistent increasing trend from 2014 to 2023.

@@

### 2.4 Relationship between Premise Types and Mental Health

@@

As illustrated in @fig-premise_type, the distribution of the number of MHA apprehensions across various premise types is presented as percentages. It can be seen that "Apartment" emerges as the predominant category in premise types. In total, 40.9% of MHA apprehension of individuals live in "Apartment". This might be due to the fact that many people are renting apartments alone or still stranger. Therefore, they have very few people to talk to if they do not feel mentally well. MHA apprehensions of individuals living in "House" has the second largest percentage (17.8%) compared to others. The number of MHA apprehensions of individuals living in "Outside" and "Other" are approximately the same, which are 16.4% and 16.7% respectively. The remaining 8.2% includes individuals living in "Commercial", "Transit", "Educational" buildings, which do not have significant effects in the context of mental health situation analysis.

@@

## 3 Reference
